Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2023-12-19 paper/paper-new.tex
updating

In this section we introduce some notations for the cumulative model that is used through the paper and in the R code. The notation is as consistent as possible with the main textbooks and with the \texttt{ordinal} R package used in the tutorial. We define \(Y_k\) as the observed ordinal response with \(1, 2, \dots,k\) options and \(Y^\star\) is the corresponding latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \alpha_2, \dots, \alpha_{k - 1}\). With \(g()\) we define the cumulative probability function assumed for the specific model. For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(f() = \Phi()\). With \(g^{-1}()\) we define the inverse of the cumulative probability function. The minus sign in \(\mathbf{X}\mathbf{\beta}\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).

\begin{equation} 
P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
\label{eq:prob-cum-model1}
\end{equation}

We can call the \(\mathbf{X}\mathbf{\beta}\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).

@@

Where \(\epsilon\) is the error part and comes from the distribution that is assumed from the latent model. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty\).

@@

\subsection{lm on latent vs ordinal}\label{lm-on-latent-vs-ordinal}

as suggested \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf}, running a lm on the latent variables gives similar parameter as the clm. Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable. But in simulation this is useful to understand what the cumulative model is doing.

@@

\begin{verbatim}
## 
## Call:
## lm(formula = ys ~ x1 + x2, data = dat)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -4.4772 -0.6765 -0.0005  0.6742  4.0796 
## 
## Coefficients:
##              Estimate Std. Error t value Pr(>|t|)    
## (Intercept) 0.0007926  0.0031780   0.249    0.803    
## x1          1.0009911  0.0031859 314.192   <2e-16 ***
## x2          0.4998991  0.0031800 157.203   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 1.005 on 99997 degrees of freedom
## Multiple R-squared:  0.5514, Adjusted R-squared:  0.5513 
## F-statistic: 6.144e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
\end{verbatim}

\begin{verbatim}
## formula: y ~ x1 + x2
## data:    dat
## 
##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
##  probit flexible  1e+05 -122909.88 245831.76 5(0)  2.04e-07 9.1e+00
## 
## Coefficients:
##     x1     x2 
## 0.9949 0.4965 
## 
## Threshold coefficients:
##     1|2     2|3     3|4     4|5 
## -0.8360 -0.2568  0.2515  0.8421
\end{verbatim}

\begin{verbatim}
##  (Intercept)           x1           x2 
## 0.0007886378 0.9960398259 0.4974264662
\end{verbatim}

\subsection{Kruscke parametrization}\label{kruscke-parametrization}

Liddell and Kruschke (2018) and Kruschke (2015) proposed an alternative parametrization to understand the model parameters. They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients. They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.

The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation). Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion. The main improvement regards mapping the values (for thresholds \(\alpha_i\) and slopes \(\beta\)) from latent standard distribution (gaussian or logistic) into the scale of the \(y\) ordinal value. The scale of the variable depend on the numeric labels assigned to ordered categories. There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to \(\alpha_1 + 0.5\) and \(\alpha_k - 0.5\) where other thresholds are estimated.

Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.

Using the function proposed by Kruschke (2015) (metti ref ad osf) and the equations presented in Kurz (2023) the \texttt{clm\_to\_ord()} function convert parameters fitted with the \texttt{clm} function into the corresponding parameters for the latent \(Y^\star\) variable considering the actual range of values (from 1 to \(k\)).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{clm\_to\_ord }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(fit)\{}
\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(fit}\SpecialCharTok{$}\NormalTok{alpha) }\CommentTok{\# estimated thresholds}
\NormalTok{  coefs }\OtherTok{\textless{}{-}}\NormalTok{ fit}\SpecialCharTok{$}\NormalTok{coefficients}
\NormalTok{  coefs }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(coefs[(}\FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(coefs)])}
\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of levels for y}
\NormalTok{  y }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k }\CommentTok{\# ordinal values}
\NormalTok{  stdm }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# latent mean}
\NormalTok{  stds }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# latent sigma}
\NormalTok{  th\_y }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(th, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{+} \FloatTok{0.5}\NormalTok{, y[k] }\SpecialCharTok{{-}} \FloatTok{0.5}\NormalTok{))}
\NormalTok{  lats }\OtherTok{\textless{}{-}}\NormalTok{ (th\_y[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th\_y[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\SpecialCharTok{/}\NormalTok{ (th[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\CommentTok{\# real latent sigma}
\NormalTok{  b0 }\OtherTok{\textless{}{-}}\NormalTok{ th\_y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ lats}
\NormalTok{  betas }\OtherTok{\textless{}{-}}\NormalTok{ lats }\SpecialCharTok{*}\NormalTok{ coefs}
\NormalTok{  out }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{sigma =}\NormalTok{ lats, }\AttributeTok{b0 =}\NormalTok{ b0, }\AttributeTok{beta =}\NormalTok{ betas, }\AttributeTok{alpha =}\NormalTok{ th\_y)}
  \FunctionTok{return}\NormalTok{(out)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{x =} \FunctionTok{rnorm}\NormalTok{(}\FloatTok{1e5}\NormalTok{)}
\NormalTok{)}

\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
\FunctionTok{clm\_to\_ord}\NormalTok{(fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $sigma
## [1] 1.78472
## 
## $b0
## [1] 2.993037
## 
## $beta
## [1] 1.797978
## 
## $alpha
## [1] 1.500000 2.560652 3.446266 4.500000
\end{verbatim}

@@

\subsection{Gelman parametrization}\label{gelman-parametrization}

rivedi questa parametrizzazione in Gelman, Hill, and Vehtari (2020)

@@

@@

@@

\subsection{Odds and odds ratio}\label{odds-and-odds-ratio}

una piccola spiegazione di odds e odds ratio

@@

Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the proportional odds assumption. There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds} models) and partially (\emph{partial proportional odds} models) (rivedi se la nomenclatura va bene). The proportional odds assumption is a crucial point when intepreting model parameters. Can be formalized as in Equation (). Basically if we use the \emph{logit} link function \(g()\) the \(\beta\)s are interpreted as standard odds ratio in the logistic regression. Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations (as reported in Equation ()). The porportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level. Thus \(\beta_1 = \beta_2 \dots = \beta_{k - 1}\). The Figure () depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).

@@

\subsection{Demonstrating the proportional odds assumption}\label{demonstrating-the-proportional-odds-assumption}

basically given that the effect of \(\beta\) is constant the odds ratio is independent from the thresholds \(\alpha_j\) (Liu et al., 2023). In the current paper we are simulating data using a single set of \(\beta\)s thus the proportional odds assumption is true. Liu et al. (2023) reviewed the available methods to test this assumption.

\begin{itemize}
\tightlist
\item
  \url{https://hbiostat.org/ordinal/impactpo.pdf} blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
\end{itemize}

sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.

Let's simulate the effect of a binary predictor on ordinal scale 1-5:

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{b1 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{3}\NormalTok{) }\CommentTok{\# log odds ratio}
\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e4}
\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
\NormalTok{probs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{) }\CommentTok{\# for the group "a", uniform probabilities}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =}\NormalTok{ b1, }\AttributeTok{probs =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}
\NormalTok{pr}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            1         2         3         4         5
## 1 0.20141151 0.2052163 0.2038036 0.1939345 0.1956341
## 2 0.07031792 0.1001600 0.1492126 0.2324907 0.4478188
\end{verbatim}

\normalsize
Basically the proportional odds suggest that:

\[
\text log(\frac{P(y \leq 1)}{P(y > 1)})
\]

Is the same regardless the level of the \(x\) predictor. Thus:

\[
\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
\]

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\begin{verbatim}
##         xa         xb 
## -0.9995719 -0.9995719
\end{verbatim}

@@

Proportional odds can be also visualized by plotting the cumulative probabilities of \(y\), in terms of \(g(P(y \leq j))\) (where \(g()\) is the logit link function) as a function of the predictor \(x\). If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption). The Figure \ref{fig:poa-plot-example} depicts the assumption of proportional odds in the probability and logit space.

>>>>

In this section we introduce some notations for the cumulative model that is used through the paper and in the R code. The notation is as consistent as possible with the main textbooks and with the \texttt{ordinal} R package used in the tutorial. We define \(Y_k\) as the observed ordinal response with \(1, 2, \dots,k\) options and \(Y^\star\) is the corresponding latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \alpha_2, \dots, \alpha_{k - 1}\). With \(g()\) we define the cumulative probability function assumed for the specific model. For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(g() = \Phi()\). With \(g^{-1}()\) we define the inverse of the cumulative probability function. The minus sign in \(\mathbf{X} \beta\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).

\begin{equation} 
P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, \dots, k - 1
\label{eq:prob-cum-model1}
\end{equation}

We can call the \(\mathbf{X} \beta\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).

@@

Where \(\epsilon\) is the error part and comes from the distribution that is assumed from the latent model. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

@@

\subsection{Alternative parametrizations}\label{alternative-parametrizations}

Several parametrizations of the ordinal models have been proposed by different authors. Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian probit model where the latent mean and standard deviations are estimated along with the threshold. The interesting part is that the thresholds are transformed to be on the scale of the variable (e.g, from 1 to 4 when \(k = 4\)) and not in terms of \(z\) scores. Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.

Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds and.

vedi qualcosa in più qui

\section{Model fitting}\label{model-fitting}

For fitting the cumulative models we used the \texttt{ordinal} package (Christensen, 2019). Despite the presence of other possibilities the \texttt{ordinal} package provide the most complete and intuitive way to implement the ordinal models. The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}.

The function to fit the model is \texttt{clm()} and the model equation is specified using the R formula syntax \texttt{y\ \textasciitilde{}\ x} where \texttt{y} is the ordinal dependent variable and \texttt{x} is one or more predictors eventually including also the interactions. The package also implements mixed-effects models (see the \texttt{clmm()} function) including random intercepts and slopes but the topic is beyond the scope of the current tutorial.

When fitting the model the crucial arguments are the \texttt{formula}, the \texttt{link} function and the \texttt{data}. More advanced arguments are the \texttt{nominal}, \texttt{scale} and \texttt{threshold}.

\begin{itemize}
\tightlist
\item
  \texttt{formula}: the formula \texttt{y\ \textasciitilde{}\ x} with the dependent variable and predictors.
\item
  \texttt{link}: is the link function. In this tutorial we consider only the \emph{logit} and \emph{probit} link but other link functions are available.
\item
  \texttt{data}: is the dataset where with the variables included in the \texttt{formula}
\item
  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section ) is relaxed (i.e., partial or non proportional odds)
\item
  \texttt{scale}: formula with predictors for the scale (standard deviation) parameter. This argument allow to fit a scale-location model (see Section ). The main \texttt{formula} argument refers to predictors on the location parameter (i.e., the mean \(\mu\)).
\item
  \texttt{threshold}: different structures for estimating the thresholds. The default is \texttt{threshold\ =\ "flexible"} where \(k - 1\) threshold (where \(k\) is the number of ordinal levels for \(Y\)) are estimated.
\end{itemize}

We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections. Table contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options. The participants are divided into two groups (predictor \(x\)). We can fit a cumulative link model with \texttt{clm()} function and check the model summary.

@@

\begin{table}

\caption{\label{tab:tab-dataset-example1}(ref:tab-dataset-example1)}
\centering
\begin{tabular}[t]{ccc}
\toprule
id & x & y\\
\midrule
1 & a & 4\\
2 & a & 2\\
3 & a & 1\\
4 & a & 3\\
... & ... & ...\\
\addlinespace
97 & b & 1\\
98 & b & 3\\
99 & b & 4\\
100 & b & 4\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{verbatim}
## $mean
##    a    b 
## 2.32 3.18 
## 
## $median
## a b 
## 2 3 
## 
## $sd
##         a         b 
## 1.0961901 0.9833305
\end{verbatim}

\begin{verbatim}
##    
##        a    b
##   1 0.30 0.10
##   2 0.26 0.10
##   3 0.26 0.32
##   4 0.18 0.48
\end{verbatim}

@@

@@

\begin{verbatim}
## formula: y ~ x
## data:    dat
## 
##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
##  logit flexible  100  -127.63 263.27 4(0)  1.25e-08 1.6e+01
## 
## Coefficients:
##    Estimate Std. Error z value Pr(>|z|)    
## xb   1.4974     0.3885   3.855 0.000116 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Threshold coefficients:
##     Estimate Std. Error z value
## 1|2  -0.8124     0.2891  -2.810
## 2|3   0.1867     0.2724   0.685
## 3|4   1.5612     0.3206   4.869
\end{verbatim}

@@

The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation. Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect. As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the intepretation depends on the contrast coding (for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, \& Kliegl, 2020)\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables. By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}

@@

\subsection{\texorpdfstring{\emph{Logit Model}. Odds and odds ratio}{Logit Model. Odds and odds ratio}}\label{logit-model.-odds-and-odds-ratio}

aggiungere cosa significano i parametri nella regressione

The odds of a probability \(p\) is defined as \(\frac{p}{1 - p}\) thus the probability of success divided by the probability of failure. The odds takes value ranging from 0 to \(\infty\). For example with a probability of \(p = 0.8\) we have an odds of \(4\), thus we have a 4 successes for each failure. The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success. When comparing two groups or conditions we can compare the two odds calculating an odds ratio. The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome. An odds ratio of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator. The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds. The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(\frac{0.5}{1 - 0.5}) = 0\). The standard logistic regression with two outcome model the logit transformed probability.

{\centering \includegraphics{paper-new_files/figure-latex/fig-odds-example-1}

\caption{(ref:fig-odds-example)}\label{fig:fig-odds-example}
\end{figure}

Odds and odds ratios are clearly defined with \(k = 2\) outcomes. With an ordinal variable and a cumulative model we can use the cumulative odds ratio. With e.g.~\(k = 4\) outcomes we have \(k - 1\) models determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\).

@@

Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the proportional odds assumption. There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds} models) and partially (\emph{partial proportional odds} models) (rivedi se la nomenclatura va bene). The proportional odds assumption is a crucial point when interpreting model parameters. Can be formalized as in Equation (). Basically if we use the \emph{logit} link function \(g()\) the \(\beta\)s are interpreted as standard odds ratio in the logistic regression. Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations (as reported in Equation ()). The porportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level. Thus \(\beta_1 = \beta_2 \dots = \beta_{k - 1}\). The Figure () depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).

@@

To make a practical example, we can use again the the dataset presented in Table. We can check the proportional odds by extrating fitted probabilities and manually computing the odds ratios.

Equation \eqref{eq:prop-odds} formalize the proportional odds assumption. Basically the cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific \(k\) or threshold \(\alpha\).

\begin{equation}
\text{logit} (\frac{P(Y \leq 1 |x_1)}{P(Y \leq 1 |x_0)}) = \dots = \text{logit} (\frac{P(Y \leq k - 1 |x_1)}{P(Y \leq k -1 |x_0)})
\label{eq:prop-odds}
\end{equation}

This can be shown by fitting the model with \texttt{clm()} that by default assume the proportional odds and using the predicted probabilities to compute the odds ratios. In the following code we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\). The log odds ratio is the same.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting the model}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}

\CommentTok{\# extracting the predicted probabilities for the two groups}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}

\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p) }\CommentTok{\# function to compute odds}

\CommentTok{\# group a}
\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\CommentTok{\# group b}
\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\begin{verbatim}
##         xa         xb 
## -0.9991102 -0.9991102
\end{verbatim}

Clearly, the odds ratio here is exactly the same because the model assume the proportional odds assumption putting this constraint on the predicted probabilities. The estimated \(\beta\) for each \(k - 1\) equation could differ and the amount of heterogeneity is actually a way to test the proportional odds with real data. In the supplementary materials we implemented an intuitive plot proposed by Harrell (2015) to visually assess the proportional odds assumption.

\subsubsection{\texorpdfstring{Proportional odds and \emph{probit} model}{Proportional odds and probit model}}\label{proportional-odds-and-probit-model}

The proportional odds assumption is relevant only for the \emph{logit} model because of the link function that put \(\beta\) as log odds. However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model. Equation \eqref{eq:probit-prop-odds} depict the parallel slopes assumption for a probit model. Basically the difference in \(z\) for a unit increase in \(x\) (i.e., the slope) is the same regardless the \(Y\) level. This can be directly demonstrated in the same way as for the logit model. We use the same dataset presented in Table \ref{tab:tab-prop-ord-ex} but fitting a probit model. The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.

\[
z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
\]

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting the model}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}

\CommentTok{\# extracting the predicted probabilities for the two groups}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}

\CommentTok{\# group a}
\NormalTok{a\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{a\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}

\CommentTok{\# group b}
\NormalTok{b\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{b\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}

\FunctionTok{c}\NormalTok{(}\AttributeTok{xa =}\NormalTok{ a\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ a\_12vs345, }\AttributeTok{xb =}\NormalTok{ b\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ b\_12vs345)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##         xa         xb 
## -0.5888457 -0.5888457
\end{verbatim}

@@

\subsubsection{Beyond the proportional odds}\label{beyond-the-proportional-odds}

qualcosa che spieghi come andare oltre e qualche proposta per testare l'assunzione.

\subsection{\texorpdfstring{\emph{Probit Model} \(z\) scores}{Probit Model z scores}}\label{probit-model-z-scores}
